---
title: "Celtic Sea Operational Model"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)
# core packages
suppressPackageStartupMessages({
  library(mizer)
  library(mizerExperimental)  
  library(mizerEcopath)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(yaml)
  library(remotes)
  library(here)
})

# ---- Local package setup ----
# We use a locally edited copy of mizerEcopath (0.1.0.9013) with master mizer (>= 2.5.3.9003).
# The master branch does not export age_mat_vB(), so we provide a small local
# compatibility wrapper (age_mat_vB_compat) and a matching newAllometricParams_compat().

# To load the local mizerEcopath source directly, use:
# devtools::load_all("C:/Documents/York academic/Projects/R projects/mizerEcopath")

```

# Introduction

This document builds the **Celtic Sea Operational Model** in mizer using the **current ** (13.10.2025) main branch (no diffusion). It is set up for scenario projections and interactive exploration (e.g. with a Shiny front end) with an emphasis on reproducibility: all inputs, choices, and manual adjustments are recorded.

**Model scope & key choices**

Fishing data: landings-only gear constructed from updated landings-by-length (scaled to Ecopath totals). Survey gear is not included as separate gear choices are not available for tuning in this mizer release.

Calibration: matchGrowth(), steadySingleSpecies(), matchBiomasses(), and matchCatch() available in this mizer release. Age-at-length data are **not** used in this release - growth is (more loosely) matched to growth curves.

Diet: we import an updated and reformatted Ecopath diet matrix.

# Modified helper functions (for newAllometricParams)

Because the master (non-diffusion) build of mizer does not export the age_mat_vB() function required by newAllometricParams(), a small local compatibility function is added here. It reproduces the expected behaviour for estimating age at maturity from VB parameters and provides a matching newAllometricParams_compat() call that mirrors the original function. This ensures the model can be built on the diffusion-free mizer version without altering the wider workflow.

```{r}
# Local compatibility version mirroring mizer::age_mat_vB()
age_mat_vB_compat <- function(object) {
  ns <- try(getNamespace("mizer"), silent = TRUE)
  if (!inherits(ns, "try-error") && exists("age_mat_vB", envir = ns, inherits = FALSE)) {
    return(get("age_mat_vB", envir = ns, inherits = FALSE)(object))
  }

  if (inherits(object, "MizerParams")) {
    sp <- object@species_params
  } else {
    if (!is.data.frame(object)) {
      stop("The first argument must be either a MizerParams object or a species_params data frame.")
    }
    sp <- validSpeciesParams(object)
  }

  sp <- set_species_param_default(sp, "t0",   0)
  sp <- set_species_param_default(sp, "b",    3)
  sp <- set_species_param_default(sp, "k_vb", NA)
  sp <- set_species_param_default(sp, "w_inf", sp$w_max)

  a_mat <- -log(1 - (sp$w_mat / sp$w_inf)^(1 / sp$b)) / sp$k_vb + sp$t0
  names(a_mat) <- sp$species
  a_mat
}


newAllometricParams_compat <- function(species_params, no_w = 200) {
  sp <- validGivenSpeciesParams(species_params)
  sp <- set_species_param_default(sp, "n", 0.7)
  sp$d <- sp$n - 1
  sp <- set_species_param_default(sp, "alpha", 0.8)
  sp$ks <- 0
  sp$z0 <- 0

  max_w <- max(sp$w_max)
  p <- newMultispeciesParams(
    sp, no_w = no_w, info_level = 0,
    max_w = max_w, w_pp_cutoff = max_w * (1 + 1e-9),
    resource_dynamics = "resource_constant"
  )
  sp <- p@species_params

  interaction_matrix(p)[] <- 0
  sp$interaction_resource <- 0
  sp$h <- Inf
  intake_max(p)[] <- Inf
  species_params(p) <- sp

  ext_encounter(p) <- t(outer(p@w, sp$n, "^"))
  sp <- set_species_param_default(sp, "age_mat", age_mat_vB_compat(sp))
  factor <- age_mat(p) / sp$age_mat
  ext_encounter(p) <- sweep(ext_encounter(p), 1, factor, "*")
  e0 <- getEGrowth(p)[, 1] / w(p)[1]^sp$n[1]

  mu0 <- e0 * (1 + 0.2 - sp$n)
  ext_mort(p) <- sweep(t(outer(p@w, sp$d, "^")), 1, mu0, "*")

  p <- matchBiomasses(p)
  p <- steadySingleSpecies(p, keep = "biomass")
  p <- setBevertonHolt(p, reproduction_level = 0)
  p
}

isAllometric_compat <- function(params, tol = 1e-5) {
  sp <- params@species_params
  sp <- set_species_param_default(sp, "d", sp$n - 1)
  m <- getMort(params) - getFMort(params)
  e <- getEncounter(params)
  for (sp_name in sp$species) {
    spc <- sp[sp_name, ]
    mc <- m[sp_name, ] / params@w^spc$d
    if (anyNA(mc) || any(is.nan(mc)) || any(mc < 0) || all(mc == 0)) stop("The model has invalid mortality rates.")
    if (abs((max(mc) - min(mc)) / min(mc)) > tol) return(FALSE)
    ec <- e[sp_name, ] / params@w^spc$n
    if (anyNA(ec) || any(is.nan(ec)) || any(ec < 0) || all(ec == 0)) stop("The model has invalid encounter rates.")
    if (abs((max(ec) - min(ec)) / min(ec)) > tol) return(FALSE)
  }
  TRUE
}
```

# Load Data

Load in the csv and rds files which contain species and empirical data. This includes data from DATRAS, STECF, and data from an existing Ecopath model (Lauria et al. 2016, including ongoing updates under Bentley et al.).

```{r}
# ---- Check for the required files
req_files <- c(
  here("inst","extdata","Ecopath-Basic estimates.csv"),
  here("inst","extdata","Ecopath-Fishing mortality rates.csv"),
  here("inst","extdata","cs_scaled_landings.csv"),
  here("inst","extdata","life_histories.csv")
)
stopifnot(all(file.exists(req_files)))

# ---- Ecopath basic estimates ----
ecopath_params <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

# ---- Ecopath fishing mortality 
yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"

# Sum all fleet columns
fleet_cols <- setdiff(seq_len(ncol(yield)), 1:2)
yield$Total <- rowSums(yield[, fleet_cols, drop = FALSE], na.rm = TRUE)

# Join to basic estimates to compute total catch (t/km^2/yr) = Total * Biomass
yield <- left_join(ecopath_params, yield, by = "Group.name") |>
  mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
  select(Group.name, TotalCatch..t.km..year.)

# ---- Landings (scaled; landings-only workflow) ----
# Created in inst/Scaling Between Observed Landings and Ecopath.qmd
cs_scaled_landings <- read.csv(here("inst","extdata","cs_scaled_landings.csv"))

# Keep only total-gear rows; bin to 1 cm; aggregate
cs_scaled_landings <- cs_scaled_landings[cs_scaled_landings$gear == "total", ] |>
  mutate(length = floor(length)) |>
  group_by(species, gear, length, Scientific_name) |>
  summarise(
    catch   = sum(scaled_catch,   na.rm = TRUE),
    biomass = sum(scaled_biomass, na.rm = TRUE),
    .groups = "drop"
  )

# Columns expected downstream
cs_scaled_landings$count <- cs_scaled_landings$catch
cs_scaled_landings$dl    <- 1L # L ensures as integer
cs_scaled_landings$gear  <- "landings"

# Reorder 
cs_scaled_landings <- cs_scaled_landings[, c(
  "species","Scientific_name","length","dl","gear","catch","count","biomass"
)]

# Combined catch table used downstream
catch <- cs_scaled_landings

# ---- Life histories (FishBase-derived; prefer Unsexed/Mixed; no size-at-age merge) ----
# (The parameter set choice is somewhat arbitrary at present)
life_histories <- read.csv(here("inst","extdata","life_histories.csv"))

life_histories_summary <- life_histories %>%
  filter(Sex %in% c("Unsexed", "Mixed")) %>%
  group_by(Species) %>%
  slice(if (any(!is.na(l_mat))) which(!is.na(l_mat))[1] else 1) %>%
  ungroup() %>%
  mutate(Scientific_name = Species)
```

# 1 Building the untuned model
## 1.1 Species parameters

We first assemble the species parameter table. We merge life-history information (length–weight, maturity length, maximum length) with Ecopath group names and ensure consistent species naming across all inputs.

The goal is to produce a clean, merged object (sp) containing key size- and weight-related parameters for each species, which will be passed to Mizer for baseline parameterisation and later tuning.

Inputs:

life_histories_summary: (FishBase-derived: a, b, l_inf, k_vb, t0, l_max, l_mat),
ecopath_params: (group names/biomass).

Note: We no longer use processed DATRAS maturity (lm) or any size-at-age merges in this non-diffusion workflow.

```{r}
# Mapping of model "common names" to scientific names used in Lauria/Ecopath
species_map_lauria <- data.frame(
  species = c("Herring","Cod","Haddock","Whiting","Blue whiting","Hake",
              "Monkfish","Horse mackerel","Mackerel","Plaice","Megrim","Sole"),
  Scientific_name = c("Clupea harengus","Gadus morhua","Melanogrammus aeglefinus",
                      "Merlangius merlangus","Micromesistius poutassou","Merluccius merluccius",
                      "Lophius piscatorius","Trachurus trachurus","Scomber scombrus",
                      "Pleuronectes platessa","Lepidorhombus whiffiagonis","Solea solea")
)

# Start from life_histories_summary (FishBase-derived only)
sp <- life_histories_summary %>%
    # add computed columns first
    mutate(
        SciName = Species,                # keep scientific name
        t0      = ifelse(is.na(t0), 0, t0),
        l_mat   = l_mat,

                # ---- Create l_max ----
        # Convert FL/SL to TL using La + Lb * value where available, else keep TL
        l_max_from_type = case_when(
          !is.na(l_max_type) & l_max_type == "TL" ~ l_max_fishbase,
          !is.na(l_max_type) & l_max_type %in% c("FL","SL") &
            !is.na(La) & !is.na(Lb) ~ La + Lb * l_max_fishbase,
          TRUE ~ NA_real_
        ),
        # NB: life_histories_summary may NOT have an existing `l_max` column,
        # so don't reference it; calculate a new one and expose later.
        l_max_calc = coalesce(l_max_from_type, l_max_fishbase),

        # ensure w_inf is present; if missing, derive from L∞ and a,b
        w_inf   = ifelse(!is.na(w_inf), w_inf,
                         ifelse(!is.na(l_inf) & !is.na(a) & !is.na(b),
                                a * (l_inf ^ b), NA_real_)),

        w_mat   = a * (l_mat ^ b),       # weight at maturity
        w_max   = a * (l_max_calc ^ b),  # max weight (from FishBase l_max)
        n       = 0.7,
        p       = 0.7,
        d       = n - 1,
        alpha   = 0.8
    ) %>%
    # join common names
    left_join(
        species_map_lauria,
        by = c("SciName" = "Scientific_name")
    ) %>%
    # overwrite species with common name
    mutate(species = species) %>%
    # keep desired columns
    transmute(
        species,     # common name 
        SciName,     # scientific name 
        w_inf,
        k_vb,
        l_mat,
        l_max = l_max_calc,
        w_max,
        w_mat,
        a, b, t0,
        n, p, d, alpha
    )

# Age at maturity from VB (needs w_inf, k_vb, t0)
sp$age_mat <- age_mat_vB_compat(sp)

```

## 1.2 Add Ecopath Parameters

Next, we link each modeled species to its corresponding Ecopath group(s) (stanza) and import biomass, production, and consumption parameters. Multi-stanza groups (e.g., Cod and Hake) are aggregated.

```{r}
species_to_groups <- list(
    Herring          = "Herring",
    Cod              = c("Cod", "Juvenile cod"),
    Haddock          = c("Haddock","Juvenile haddock"),
    Whiting          = c("Whiting","Juvenile whiting"),
    `Blue whiting`   = c("Blue whiting","Juvenile blue whiting"),
    Hake             = c("Hake","Juvenile hake"),
    Monkfish         = c("Monkfish","Juvenile monkfish"),
    `Horse mackerel` = "Horse mackerel",
    Mackerel         = "Mackerel",
    Plaice           = c("Plaice","Juvenile plaice"),
    Megrim           = c("Megrim","Juvenile megrim"),
    Sole             = "Sole"
)

# Use addEcopathParams to validate and add parameters from Ecopath to each species.
sp <- addEcopathParams(sp, ecopath_params, species_to_groups)
```

## 1.3 Update w_max from observations 

We adjust each species’ maximum observed length (`l_max`) using the largest length recorded in the scaled landings dataset (`cs_scaled_landings`), which originates from **cs_catch.rds** (observed landings by length, STECF) scaled to match **Ecopath total catches**.

```{r}
## 1.3 Update w_max from observations 
# Use the largest observed length in the (scaled) landings and give a modest buffer.

observed_max <- cs_scaled_landings %>%
  group_by(species) %>%
  summarise(
    obs_len = max(length + dl, na.rm = TRUE),
    .groups = "drop"
  )

sp <- sp %>%
  left_join(observed_max, by = "species") %>%
  mutate(
    # keep l_max if already larger; otherwise raise it up to the observed max
    l_max = if_else(!is.na(obs_len) & obs_len > l_max, obs_len, l_max),
    # set w_max with a 20% buffer 
    len_for_max = coalesce(obs_len, l_max),
    w_max       = 1.2 * a * (len_for_max ^ b),
    # match Lauria’s practice of setting the upper reproductive weight to w_inf
    w_repro_max = w_inf
  ) %>%
  select(-len_for_max, -obs_len)

```

## 1.4 Build non‑interacting mizer model

We next set up a 'non‑interacting' mizer model where each species grows according to allometric rules independently of each other species in the assemblage. This essentially generates a series of independent baseline parameters for each species, ready to be matched to observation data.

The process is:

1. Create allometric size spectra for each species.
2. Add Ecopath total catch targets (for catch matching).
3. Assign observed biomasses from Ecopath for matching.
4. Iteratively match growth, biomass, and consumption rates.

```{r}
# Set up independent (non-interacting) species spectra from sp
model_non_int <- newAllometricParams_compat(sp, no_w = 200)

# Attach Ecopath-derived total catch targets used by matching
# We created `yield` above with TotalCatch..t.km..year.; convert to expected column names
catch_df <- yield %>%
  transmute(
    `Group name` = Group.name,
    `TotalCatch (t/km²/year)` = TotalCatch..t.km..year.
  )

model_non_int <- addEcopathCatchTotal(model_non_int, catch_df)

# ---- Match growth & biomass to observations ----
# Map each species to its adult stanza and join adult biomass from Ecopath.
# We collapse Ecopath to one row per group name to stay robust if the file contains duplicates.
adult_map <- tibble::tibble(
  species      = names(species_to_groups),
  adult_stanza = vapply(species_to_groups, function(x) x[[1]], character(1))
)

adult_biomass <- ecopath_params %>%
  group_by(Group.name) %>%
  summarise(
    biomass_observed = sum(Biomass..t.km.2., na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(adult_stanza = Group.name)

species_params_tbl <- species_params(model_non_int) %>%
  select(-any_of("biomass_observed")) %>%  # drop if present from a previous run
  left_join(adult_map, by = "species") %>%
  left_join(adult_biomass, by = "adult_stanza") %>%
  mutate(
    biomass_observed = coalesce(biomass_observed, 0),
    biomass_cutoff   = w_mat
  ) %>%
  select(-adult_stanza)

species_params(model_non_int) <- species_params_tbl

# Calibration to single-species targets
model_non_int <- model_non_int %>%
  matchGrowth() %>%
  steadySingleSpecies() %>%
  matchBiomasses()


```
## 1.5 Match catch and consumption (landings‑only, not survey)

We now match catches to the landings‑only observations and then align consumption with Ecopath reference values. (Survey and size‑at‑age inputs are not used in this non‑diffusion build.)

```{r}
## 1.5 Match catch and consumption (landings-only)
# - use landings only
# - standardise a single gear label

landings_total <- cs_scaled_landings %>%
  dplyr::mutate(
    # harmonise casing to model names if needed
    species = dplyr::recode(species, "Horse Mackerel" = "Horse mackerel")
  ) %>%
  dplyr::select(species, length, dl, count) %>%
  dplyr::mutate(gear = "total")

model_non_int <- model_non_int %>%
  matchCatch(catch = landings_total, yield_lambda = 1) %>%
  matchConsumption()

```

## 1.6 Add diet and prepare for interactions.
In this step we add diet information to the model, using the updated Ecopath diet matrix (reformatted species × prey table). This enables diet‑based tuning and subsequent activation of  interactions with (matchDiet()).

```{r}
## 1.6 Add diet and prepare for interactions --------------------------------
## Reduce the Ecopath diet to a species×species matrix and set kernels from
## stomach-fit summaries (no joins; align by order like the original CS build).

# Diet (Ecopath -> model species × species)
ecopath_diet <- read.csv(
  here("inst", "extdata", "Ecopath-Diet_updated_reformatted.csv"),
  check.names = FALSE
)

dm <- reduceEcopathDiet(
  species_params = species_params(model_non_int),
  ecopath_diet   = ecopath_diet,
  ecopath_params = ecopath_params
)

# Stomach-fit parameters (keep separate; align by scientific name)
stf <- readRDS(here("inst", "extdata", "stomach_data_fit.rds"))

sp  <- species_params(model_non_int)
idx <- match(sp$SciName, stf$species)  # align order to species_params

lambda <- 2
sp$pred_kernel_type <- "power_law"
sp$kernel_exp <- stf$alpha[idx] + 4/3 - lambda
sp$kernel_l_l <- stf$ll[idx]
sp$kernel_u_l <- stf$ul[idx]
sp$kernel_l_r <- stf$lr[idx]
sp$kernel_u_r <- stf$ur[idx]

species_params(model_non_int) <- sp

# Save a checkpoint
#saveRDS(model_non_int, here("data","model_non_int.rds"))
```

## 1.7 Pre-tuning set up
Sets up elements required before calling tuneEcopath

```{r}
# Directory for checkpoints
chk_dir <- here("data")

# Modify landings data for tuning (must have species, length, dl, catch, gear)
catch_for_tune <- cs_scaled_landings %>%
  mutate(
    species = recode(species, "Horse Mackerel" = "Horse mackerel"),
    gear    = "total"
  ) %>%
  transmute(
    species,
    length,
    dl     = as.integer(dl),
    catch  = as.numeric(count),
    gear
  )

# Empty survey age-at-length stub so the gadget's Survey Age tab does not error [temporary solution due to using new mizerEcopath gadget]
age_at_length_empty <- tibble(
  species = character(),
  Scientific_name = character(),
  survey  = character(),
  length  = numeric(),
  dl      = integer(),
  age     = integer(),
  count   = numeric()
)
```

# 2 Manual tuning 

Rounds of re-tuning with saved model object. 

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Load the checkpoint saved at the end of 1.6 if required
#model_non_int <- readRDS(file.path(chk_dir, "model_non_int.rds"))

# Run the tuner - diffusion sliders will appear but have no effect on the model.  Reuse model_non_int_1 as the base object if partially tuned.
model_non_int_1 <- tuneEcopath(
    model_non_int,
    controls = c("match","growth",
                 "diffusion",
                 "fishing",
                 "reproduction",
                 "other"),
    catch         = catch_for_tune,
    age_at_length = age_at_length_empty,
    diet          = dm
)

# With the set up above, untick catch, press match, to save a state which can be returned to (pressing with catch ticked may return to the original catch configuration, which controls catchability, selectivity, and mortality).

# Optional: save tuned checkpoint [will need to update the tuneEcopath command to use this instead to continue]
# saveRDS(model_non_int_1, file.path(chk_dir, "model_tuned_1.rds"))

```

## 2.1 Further tuning on existing model object

Use this to reopen tuneEcopath on the current model (model_non_int_1) without rebuilding. We explicitly set controls so the UI initialises cleanly (the diffusion tab is shown but inert). The app is fed the scaled landings (catch_for_tune), an empty age–length placeholder (age_at_length_empty), and the Ecopath-reduced diet matrix (dm).

```{r, warning=FALSE, message=FALSE, error=FALSE}
#model_non_int_1 <- readRDS(file.path(chk_dir, "model_tuned_1.rds"))

#Area to rework partially tuned model
model_non_int_1 <- tuneEcopath(
    model_non_int_1,
    controls = c("match","growth",
                 "diffusion",
                 "fishing",
                 "reproduction",
                 "other"),
    catch         = catch_for_tune,
    age_at_length = age_at_length_empty,
    diet          = dm
)

# saveRDS(model_non_int_1, file.path(chk_dir, "model_tuned_1.rds"))
```

# Post-hoc helpers
Sections below contain helpers which can be used to make changes to the feeding kernels and observed biomasses of species. May help if the size distribution curtails too early in the model vs observed.

## Post hoc: Asymptotic size adjustment

```{r}
model_non_int_1@species_params$species
#model_non_int_1@species_params$w_repro_max[6] <- 1400 
#model_non_int_1@species_params$w_repro_max[3] <- 1250 
#model_non_int_1@species_params$w_repro_max[11] <- 1000 
#model_non_int_1 <- setReproduction(model_non_int_1)
```

## Post-hoc: Kernel adjustment

Use this section to edits to the predator–prey kernels.

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Choose model
p <- model_non_int_1 

# Locate species row
i <- match("Megrim", species_params(p)$species)

# Backup current values so they can be reverted easily
sp <- species_params(p)
backup_hake <- sp[i, c("kernel_l_l","kernel_u_l","kernel_l_r","kernel_u_r","kernel_exp")]
print(backup_hake)

# Show current cut-offs on PPMR (mass ratio) scale for intuition (value is factor bigger of the predator than the prey).
cat("Current cut-offs on PPMR scale:\n")
cat("  lower (start-eating)  ~", exp(sp$kernel_l_l[i]), "\n")
cat("  upper (stop-eating)   ~", exp(sp$kernel_l_r[i]), "\n")

# --- Move Hake's lower cutoff to prey ≈ 1/10 of predator mass (PPMR = 10) ---
# (Previously around 80x: exp(sp$kernel_l_l[i]) ~ 80)
#sp$kernel_l_l[i] <- log(10)   # ≈ 2.302585; shifts rising sigmoid right to PPMR=10

# --- Move Megrim's lower cutoff to prey ≈ 1/30 of predator mass  ---
# (Previously around 50x)
#sp$kernel_l_l[i] <- log(30)   

# Write back and rebuild kernel lookups
species_params(p) <- sp
p <- setPredKernel(p)

# Inspect new PPMR cut-offs
cat("New cut-offs on PPMR scale:\n")
cat("  lower ~", exp(species_params(p)$kernel_l_l[i]), "\n")
cat("  upper ~", exp(species_params(p)$kernel_l_r[i]), "\n")

# Finally put into the real model
model_non_int_1 <- p
```

## Post-hoc: Biomass adjustment

This helper allows quick manual scaling of the observed biomass target for selected species after tuning. It’s intended for fine adjustments — useful when Ecopath totals look implausible. The change affects reproduction equilibrium only and does not alter gear selectivity or fishing mortality. After scaling, the model re-equilibrates with matchBiomasses() and steadySingleSpecies() to maintain balance.

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Post-hoc: multiply observed biomass targets and show a simple before/after
# - factors: named numeric vector (e.g. c("Megrim" = 1.4, "Cod" = 1.2))
#            OR a 2-column data.frame: species, factor
# - reequilibrate: if TRUE, run matchBiomasses() and steadySingleSpecies()
# Returns the updated model and prints a table for changed species.
scale_observed_biomass <- function(p, factors, reequilibrate = TRUE, digits = 3) {
  sp <- species_params(p)
  sp_names <- as.character(sp$species)

  # Normalise `factors` to a named numeric vector
  if (is.data.frame(factors)) {
    if (ncol(factors) < 2) stop("Provide a 2-column data.frame: species, factor")
    fvec <- setNames(as.numeric(factors[[2]]), as.character(factors[[1]]))
  } else if (is.numeric(factors) && !is.null(names(factors))) {
    fvec <- factors
  } else {
    stop("Provide a named numeric vector (names = species) or a 2-col data.frame.")
  }

  # Keep only species that exist in the model
  hit <- intersect(names(fvec), sp_names)
  if (!length(hit)) stop("None of the provided species match species_params$species")

  # Save original targets
  original_target <- setNames(as.numeric(sp$biomass_observed), sp_names)

  # Apply multiplicative scaling to targets
  idx <- match(hit, sp_names)
  sp$biomass_observed[idx] <- sp$biomass_observed[idx] * fvec[hit]
  species_params(p) <- sp

  # Re-equilibrate if requested (by default runs matchBiomasses and steady).
  if (reequilibrate) {
    p <- matchBiomasses(p)
    p <- steadySingleSpecies(p, keep = "biomass")
  }

  # New targets + realised biomasses from the model
  sp_after <- species_params(p)
  new_target <- setNames(as.numeric(sp_after$biomass_observed), as.character(sp_after$species))

  realised_vec <- tryCatch(
    getBiomass(p),
    error = function(e) setNames(rep(NA_real_, length(sp_names)), sp_names)
  )

  # Small, readable comparison for the species you changed
  out <- data.frame(
    species             = hit,
    original_target     = as.numeric(original_target[hit]),
    new_target          = as.numeric(new_target[hit]),
    realised            = as.numeric(realised_vec[hit]),
    realised_to_target  = as.numeric(realised_vec[hit]) /
                          pmax(as.numeric(new_target[hit]), .Machine$double.eps),
    check.names = FALSE
  )

  # Pretty print (optional rounding)
  num_cols <- vapply(out, is.numeric, logical(1))
  out[num_cols] <- lapply(out[num_cols], function(x) signif(x, digits))
  rownames(out) <- NULL
  print(out)
  invisible(p)
}

    ## Example usage:
# factors <- c("Megrim" = 1.4, "Cod" = 1.2)
# model_non_int_1 <- scale_observed_biomass(model_non_int_1, factors)

####    Use biomass rescaler    ######################
factors_fish_biomass <- c("Whiting" = 2) # Other uses run in console, check with helper below
model_non_int_1 <- scale_observed_biomass(model_non_int_1, factors_fish_biomass, reequilibrate = TRUE)
######################################################
### Compare biomass in model to original to check for any changes:

compare_biomass_to_ecopath <- function(mod) {
  # Accept either an object or a character name
  if (is.character(mod)) mod <- get(mod, envir = parent.frame())

  adult_map <- data.frame(
    species      = names(species_to_groups),
    adult_stanza = vapply(species_to_groups, function(x) x[[1]], character(1))
  )

  ecopath_adult <- ecopath_params %>%
    group_by(Group.name) %>%
    summarise(bio_ecopath = sum(Biomass..t.km.2., na.rm = TRUE), .groups = "drop")

  out <- species_params(mod) %>% as.data.frame() %>%
    select(species, bio_model = biomass_observed) %>%
    left_join(adult_map, by = "species") %>%
    left_join(ecopath_adult, by = c("adult_stanza" = "Group.name")) %>%
    mutate(ratio = bio_model / pmax(bio_ecopath, .Machine$double.eps)) %>%
    arrange(desc(abs(log10(pmax(ratio, .Machine$double.eps))))) %>%
    select(species, bio_ecopath, bio_model, ratio)

  print(out)
  invisible(out)
}

# example:
compare_biomass_to_ecopath(model_non_int_1)

```

# Comments

use compareParams to see which model properties have changed between model iterations.

In calibrating the Celtic Sea operational model I prioritise matching total yield and the catch-at-length composition, because these are the quantities most directly informed by the data (fishing mortality applied to estimated biomasses and the observed length structure of landings). Once those are in good order, I focus on interspecific interactions: predators must be able to meet their consumption needs without driving prey below sustainable levels at equilibrium. In practice that means ensuring that diets and size-based feeding preferences remain meet energy needs and are ecologically plauisible.

Where strict agreement across all constraints is not possible, I allow controlled relaxation of either production or biomass assumptions. The aim is a transparent, steady-state calibration that balances fit to yield and length composition with trophic realism. Any departures from reference values will need to be recorded with a brief rationale so their implications for scenarios can be interpreted clearly later as required.

Notes - Megrim; growth rate much higher than VBG, but production similar to estimate. Minor increases in biomass (up to 1.5 x) insufficient to make much difference (it remains hard to keep diet proportion > 1). At time of writing herring only has a 1.5 x increase in biomass vs base estimate. 

# 3 Setting reproduction levels
## F_MSY for each species
Herring - 0.266 (Herring (Clupea harengus) in Division 7.a North of 52°30’N (Irish Sea), 2021)
Cod - 0.29 (Cod Gadus morhua in Divisions 7.e–k, Celtic Sea & Western Channel, 2021)
Megrim - 0.191 (Megrim Lepidorhombus whiffiagonis in Divisions 7.b–k, 8.a–b, and 8.d; ICES Advice 2021, Table 4 “Reference points”)
Monkfish - 0.28 (White anglerfish, Lophius piscatorius; “White anglerfish in Subarea 7 and divisions 8.a–b & 8.d (Celtic Seas, Bay of Biscay)”, ICES Advice 2021, Table 4: Reference points)
Haddock - 0.353 (Haddock (Melanogrammus aeglefinus) in divisions 7.b–k (southern Celtic Seas and English Channel), ICES Advice 2022)
Whiting - 0.375 (Whiting (Merlangius merlangus) in divisions 7.b–c and 7.e–k (southern Celtic Seas and western English Channel), ICES Advice 2022)
Hake - 0.24 (Hake Merluccius merluccius, Northern stock in subareas 4, 6, 7 and divisions 3.a, 8.a–b, 8.d; ICES Advice 2022)
Blue whiting - 0.32 (Blue whiting Micromesistius poutassou in subareas 1–9, 12, and 14; ICES Advice 2022, Table 4 “Reference points”)
Plaice - 0.196 (Plaice Pleuronectes platessa in Division 7.a (Irish Sea), ICES Advice 2021)
Mackerel - 0.26 (Mackerel Scomber scombrus in subareas 1–8 & 14 and Division 9.a (Northeast Atlantic & adjacent waters), ICES Advice 2022)
Sole - 0.251 (Sole Solea solea in divisions 7.f–g; Bristol Channel & Celtic Sea), ICES Advice 2023
Horse mackerel - 0.074 (Western horse mackerel Trachurus trachurus in Subarea 8 and Divisions 2.a, 4.a, 5.b, 6.a, 7.a–c, and 7.e–k; ICES Advice 2022)

```{r, warning=FALSE, message=FALSE, error=FALSE}
# --- Target table: place yield-curve peak at F_MSY * (q / F_ecopath_adult) ----
# 1) F_MSY values (source notes kept in prose above)
fmsy_vals <- c(
  "Herring"        = 0.266,
  "Cod"            = 0.29,
  "Megrim"         = 0.191,
  "Monkfish"       = 0.28,
  "Haddock"        = 0.353,
  "Whiting"        = 0.375,
  "Hake"           = 0.24,
  "Blue whiting"   = 0.32,
  "Plaice"         = 0.196,  # proxy (Irish Sea)
  "Mackerel"       = 0.26,
  "Sole"           = 0.251,
  "Horse mackerel" = 0.074
)
fmsy_tbl <- tibble(species = names(fmsy_vals), F_msy = unname(fmsy_vals))

# 2) Read Ecopath F file and sum across fleets to get total F per Ecopath group
ecoF <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"),
                 check.names = FALSE)

# Normalise column names (handle empty/NA headers)
nm <- names(ecoF)
empty <- which(is.na(nm) | nm == "")
if (length(empty)) nm[empty] <- paste0("unnamed_", empty)
names(ecoF) <- nm

# Identify the group-name column (first header containing "group"; else fall back to 2nd col)
grp_col <- which(grepl("group", names(ecoF), ignore.case = TRUE))[1]
if (is.na(grp_col)) grp_col <- 2
names(ecoF)[grp_col] <- "Group.name"

# Fleet columns = all except first and the group column
fleet_cols <- setdiff(seq_len(ncol(ecoF)), c(1, grp_col))
stopifnot(length(fleet_cols) > 0)

# Coerce fleet columns to numeric and sum to F_total
ecoF[fleet_cols] <- lapply(ecoF[fleet_cols], function(x) suppressWarnings(as.numeric(x)))
ecoF_total <- data.frame(
  Group.name = ecoF$Group.name,
  F_ecopath_adult = rowSums(ecoF[, fleet_cols, drop = FALSE], na.rm = TRUE),
  check.names = FALSE
)
# Drop NA group names and duplicates
ecoF_total <- ecoF_total[!is.na(ecoF_total$Group.name), , drop = FALSE]
ecoF_total <- ecoF_total[!duplicated(ecoF_total$Group.name), , drop = FALSE]

# 3) Map Ecopath F_total to the adult stanza of each model species
eco_F_adult <- adult_map |>
  left_join(ecoF_total, by = c("adult_stanza" = "Group.name")) |>
  select(species, F_ecopath_adult)

# 4) Per-species catchability q 
q_tbl <- gear_params(model_non_int_1) |>
  group_by(species) |>
  summarise(q = dplyr::first(na.omit(catchability)), .groups = "drop")

# 5) Assemble target table
# F_target_peak = F_MSY * (q / F_ecopath_adult)
targets_out <- species_params(p_int) |>
  as.data.frame() |>
  select(species) |>
  left_join(eco_F_adult, by = "species") |>
  left_join(q_tbl,        by = "species") |>
  left_join(fmsy_tbl,     by = "species") |>
  mutate(
    F_target_peak = ifelse(is.finite(F_msy * q / F_ecopath_adult),
                           F_msy * q / F_ecopath_adult, NA_real_)
  ) |>
  select(species, F_msy, q, F_ecopath_adult, F_target_peak) |>
  arrange(species)

print(targets_out)

```

```{r, warning=FALSE, message=FALSE, error=FALSE}
# ---- Yield-curve helpers (single-species) -----------------------------------

# repro_lvl_original <- getReproductionLevel(model_non_int_1)
getReproductionLevel(model_non_int_1)

# Return a scalar F_target_peak for one species (or NA if not found)
get_Ft_one <- function(species_name, targets_tbl = targets_out) {
  stopifnot(is.character(species_name), length(species_name) == 1)
  i <- match(species_name, targets_tbl$species)
  if (is.na(i)) return(NA_real_)
  as.numeric(targets_tbl$F_target_peak[i])
}

# Plot yield vs F for one species, optionally add a vertical line at target F
plot_yield_with_target <- function(p, species_name,
                                   expand = 1.5, step = 0.02, tol = 1e-4,
                                   targets_tbl = targets_out) {
  stopifnot(is.character(species_name), length(species_name) == 1)
  Ft <- get_Ft_one(species_name, targets_tbl)
  Fmax <- if (is.finite(Ft)) max(0.9, expand * Ft) else 1.0
  g <- plotYieldVsF(
    p, species = species_name,
    F_range = seq(0, Fmax, by = step),
    tol = tol
  )
  if (is.finite(Ft)) g <- g + ggplot2::geom_vline(xintercept = Ft, linetype = 2)
  g
}

# Edit reproduction level for a single species and return updated model
set_reproduction_level_one <- function(p, species_name, level) {
  stopifnot(is.character(species_name), length(species_name) == 1)
  rl <- getReproductionLevel(p)
  if (!(species_name %in% names(rl))) {
    stop("Species not found in reproduction_level vector: ", species_name)
  }
  rl[species_name] <- level
  setBevertonHolt(p, reproduction_level = rl)
}

# ---- Example usage ----------------------------------------------------------
# Species list
model_non_int_1@species_params$species

# Choose a species
sp_name <- "Herring"

# Inspect current target_F
Ft <- get_Ft_one(sp_name)   
print(Ft)

# Plot current yield curve with target marker
plot_yield_with_target(model_non_int_1, sp_name)

# Tweak reproduction level and re-plot
model_non_int_1 <- set_reproduction_level_one(model_non_int_1, sp_name, 0.215)
plot_yield_with_target(model_non_int_1, sp_name)

# Stash reproduction levels if not already saved into model:
repro_lvl_tuned <- getReproductionLevel(model_non_int_1)

```

# Resilience comments
With landings-only selectivity and current background mortality, Megrim’s yield peak stays around 0.32–0.35 1/yr, above the target 0.158 1/yr. Reproduction-level changes beyond ~1e-2 have diminishing impact. In future, could be changed by altering l_mat - would have to retune at least Megrim, and possibly other species too if unlucky (due to diet/mortality shifts). 

Second note - I tuned resilience to the non-interacting model, probably more sensible to do for the interacting version. Will use the same workflow. 

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Save the resilience-tuned non-interacting model
#saveRDS(model_non_int_1, file.path(here::here("data"), "model_non_int_resilience_tuned.rds"))

model_non_int_res <- readRDS(file.path(here::here("data"), "model_non_int_resilience_tuned.rds"))
```

# 4 Turning on interactions and resource dynamics

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Start from the non-interacting, reproduction-tuned model
p0 <- model_non_int_res

# 1) Add a bit of satiation before interactions
p0 <- setFeedingLevel(p0, 0.6)

# 2) Switch on species interactions
p_int <- matchDiet(p0, dm)

# 3) Align the resource to the fish community and make it dynamic
resource_params(p_int)$w_pp_cutoff <- 1
resource_params(p_int)$n <- 0.7
p_int <- alignResource(p_int)
p_int <- setResourceInteraction(p_int,
                                resource_dynamics = "resource_semichemostat")

plotSpectra(p_int)

# Project to steady if required
p_int <- projectToSteady(p_int)

#saveRDS(p_int, file.path(here::here("data"), "full_tuned_cs_model.rds"))
```


