---
title: "Celtic Sea Operational Model"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)
# core packages
suppressPackageStartupMessages({
  library(mizer)
  library(mizerEcopath)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(yaml)
  library(remotes)
  library(here)
})

# ---- Local package setup ----
# Load mizerEcopath from a locally edited copy rather than GitHub.
# This local version (0.1.0.9013) was installed from JEVER's fork
# of the spawning_season branch with a modified DESCRIPTION file.
# The modification relaxes the mizer dependency from >= 2.5.3.9012 to >= 2.5.3.9003
# so it runs on the current master (non-diffusion) mizer build. 
# If not loaded from local copy, use:
# devtools::load_all("C:/Documents/York academic/Projects/R projects/mizerEcopath")

```

# Introduction

This document builds the **Celtic Sea Operational Model** in mizer using the **current ** (13.10.2025) main branch (no diffusion). t is configured for policy-facing projections and interactive exploration (e.g. with a Shiny front end) with an emphasis on reproducibility: all inputs, choices, and manual adjustments are recorded.

Model scope & key choices

Fishing data: landings-only gear constructed from updated landings-by-length (scaled to Ecopath totals). Survey gear is not included as seperate gear choices are not available for tuning in this mizer release.

Calibration: matchGrowth(), steadySingleSpecies(), matchBiomasses(), and matchCatch() available in this mizer release. Age-at-length data are **not** used in this release - growth is (more loosely) matched to growth curves.

Diet: we import an updated and reformatted Ecopath diet matrix.

# Load Data

Load in the csv and rds files which contain species and empirical data. This includes data from DATRAS, STECF, and data from an existing Ecopath model (Lauria et al. 2016, including ongoing updates under Bentley et al.).

```{r}
# ---- Check for the required files
req_files <- c(
  here("inst","extdata","Ecopath-Basic estimates.csv"),
  here("inst","extdata","Ecopath-Fishing mortality rates.csv"),
  here("inst","extdata","cs_scaled_landings.csv"),
  here("inst","extdata","life_histories.csv")
)
stopifnot(all(file.exists(req_files)))

# ---- Ecopath basic estimates ----
ecopath_params <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

# ---- Ecopath fishing mortality 
yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"

# Sum all fleet columns
fleet_cols <- setdiff(seq_len(ncol(yield)), 1:2)
yield$Total <- rowSums(yield[, fleet_cols, drop = FALSE], na.rm = TRUE)

# Join to basic estimates to compute total catch (t/km^2/yr) = Total * Biomass
yield <- dplyr::left_join(ecopath_params, yield, by = "Group.name") |>
  dplyr::mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
  dplyr::select(Group.name, TotalCatch..t.km..year.)

# ---- Landings (scaled; landings-only workflow) ----
# Created in inst/Scaling Between Observed Landings and Ecopath.qmd
cs_scaled_landings <- read.csv(here("inst","extdata","cs_scaled_landings.csv"))

# Keep only total-gear rows; bin to 1 cm; aggregate
cs_scaled_landings <- cs_scaled_landings[cs_scaled_landings$gear == "total", ] |>
  dplyr::mutate(length = floor(length)) |>
  dplyr::group_by(species, gear, length, Scientific_name) |>
  dplyr::summarise(
    catch   = sum(scaled_catch,   na.rm = TRUE),
    biomass = sum(scaled_biomass, na.rm = TRUE),
    .groups = "drop"
  )

# Columns expected downstream
cs_scaled_landings$count <- cs_scaled_landings$catch
cs_scaled_landings$dl    <- 1L # L ensures as integer
cs_scaled_landings$gear  <- "landings"

# Reorder 
cs_scaled_landings <- cs_scaled_landings[, c(
  "species","Scientific_name","length","dl","gear","catch","count","biomass"
)]

# Combined catch table used downstream
catch <- cs_scaled_landings

# ---- Life histories (FishBase-derived; prefer Unsexed/Mixed; no size-at-age merge) ----
life_histories <- read.csv(here("inst","extdata","life_histories.csv"))

life_histories_summary <- life_histories %>%
  dplyr::filter(Sex %in% c("Unsexed", "Mixed")) %>%
  dplyr::group_by(Species) %>%
  dplyr::slice(if (any(!is.na(l_mat))) which(!is.na(l_mat))[1] else 1) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Scientific_name = Species)
```



