---
title: "Celtic Sea Operational Model"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)
# core packages
suppressPackageStartupMessages({
  library(mizer)
  library(mizerEcopath)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(yaml)
  library(remotes)
  library(here)
})

# ---- Local package setup ----
# Load mizerEcopath from a locally edited copy rather than GitHub.
# This local version (0.1.0.9013) was installed from JEVER's fork
# of the spawning_season branch with a modified DESCRIPTION file.
# The modification relaxes the mizer dependency from >= 2.5.3.9012 to >= 2.5.3.9003
# so it runs on the current master (non-diffusion) mizer build. 
# If not loaded from local copy, use:
# devtools::load_all("C:/Documents/York academic/Projects/R projects/mizerEcopath")

```

# Introduction

This document builds the **Celtic Sea Operational Model** in mizer using the **current ** (13.10.2025) main branch (no diffusion). t is configured for policy-facing projections and interactive exploration (e.g. with a Shiny front end) with an emphasis on reproducibility: all inputs, choices, and manual adjustments are recorded.

Model scope & key choices

Fishing data: landings-only gear constructed from updated landings-by-length (scaled to Ecopath totals). Survey gear is not included as seperate gear choices are not available for tuning in this mizer release.

Calibration: matchGrowth(), steadySingleSpecies(), matchBiomasses(), and matchCatch() available in this mizer release. Age-at-length data are **not** used in this release - growth is (more loosely) matched to growth curves.

Diet: we import an updated and reformatted Ecopath diet matrix.

# Load Data

Load in the csv and rds files which contain species and empirical data. This includes data from DATRAS, STECF, and data from an existing Ecopath model (Lauria et al. 2016, including ongoing updates under Bentley et al.).

```{r}
# ---- Check for the required files
req_files <- c(
  here("inst","extdata","Ecopath-Basic estimates.csv"),
  here("inst","extdata","Ecopath-Fishing mortality rates.csv"),
  here("inst","extdata","cs_scaled_landings.csv"),
  here("inst","extdata","life_histories.csv")
)
stopifnot(all(file.exists(req_files)))

# ---- Ecopath basic estimates ----
ecopath_params <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

# ---- Ecopath fishing mortality 
yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"

# Sum all fleet columns
fleet_cols <- setdiff(seq_len(ncol(yield)), 1:2)
yield$Total <- rowSums(yield[, fleet_cols, drop = FALSE], na.rm = TRUE)

# Join to basic estimates to compute total catch (t/km^2/yr) = Total * Biomass
yield <- dplyr::left_join(ecopath_params, yield, by = "Group.name") |>
  dplyr::mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
  dplyr::select(Group.name, TotalCatch..t.km..year.)

# ---- Landings (scaled; landings-only workflow) ----
# Created in inst/Scaling Between Observed Landings and Ecopath.qmd
cs_scaled_landings <- read.csv(here("inst","extdata","cs_scaled_landings.csv"))

# Keep only total-gear rows; bin to 1 cm; aggregate
cs_scaled_landings <- cs_scaled_landings[cs_scaled_landings$gear == "total", ] |>
  dplyr::mutate(length = floor(length)) |>
  dplyr::group_by(species, gear, length, Scientific_name) |>
  dplyr::summarise(
    catch   = sum(scaled_catch,   na.rm = TRUE),
    biomass = sum(scaled_biomass, na.rm = TRUE),
    .groups = "drop"
  )

# Columns expected downstream
cs_scaled_landings$count <- cs_scaled_landings$catch
cs_scaled_landings$dl    <- 1L # L ensures as integer
cs_scaled_landings$gear  <- "landings"

# Reorder 
cs_scaled_landings <- cs_scaled_landings[, c(
  "species","Scientific_name","length","dl","gear","catch","count","biomass"
)]

# Combined catch table used downstream
catch <- cs_scaled_landings

# ---- Life histories (FishBase-derived; prefer Unsexed/Mixed; no size-at-age merge) ----
life_histories <- read.csv(here("inst","extdata","life_histories.csv"))

life_histories_summary <- life_histories %>%
  dplyr::filter(Sex %in% c("Unsexed", "Mixed")) %>%
  dplyr::group_by(Species) %>%
  dplyr::slice(if (any(!is.na(l_mat))) which(!is.na(l_mat))[1] else 1) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Scientific_name = Species)
```

# Step 1
## 1.1 Species parameters

We first assemble the species parameter table. We merge life-history information (length–weight, maturity length, maximum length) with Ecopath group names and ensure consistent species naming across all inputs.

The goal is to produce a clean, merged object (sp) containing key size- and weight-related parameters for each species, which will be passed to Mizer for baseline parameterisation and later tuning.

Inputs:

life_histories_summary: (FishBase-derived: a, b, l_inf, k_vb, t0, l_max, l_mat),
ecopath_params: (group names/biomass).

Note: We no longer use processed DATRAS maturity (lm) or any size-at-age merges in this non-diffusion workflow.

```{r}
# Mapping of model "common names" to scientific names used in Lauria/Ecopath
species_map_lauria <- data.frame(
  species = c("Herring","Cod","Haddock","Whiting","Blue whiting","Hake",
              "Monkfish","Horse mackerel","Mackerel","Plaice","Megrim","Sole"),
  Scientific_name = c("Clupea harengus","Gadus morhua","Melanogrammus aeglefinus",
                      "Merlangius merlangus","Micromesistius poutassou","Merluccius merluccius",
                      "Lophius piscatorius","Trachurus trachurus","Scomber scombrus",
                      "Pleuronectes platessa","Lepidorhombus whiffiagonis","Solea solea")
)

# Start from life_histories_summary (FishBase-derived only; no DATRAS maturity merge in this workflow)
sp <- life_histories_summary %>%
    # add computed columns first
    mutate(
        SciName = Species,                # keep scientific name
        t0      = ifelse(is.na(t0), 0, t0),
        # use l_mat as provided (FishBase-derived); we are not overriding with DATRAS
        l_mat   = l_mat,

        # ---- Recreate l_max as in old workflow ----
        # Convert FL/SL to TL using La + Lb * value where available, else keep TL
        l_max_from_type = dplyr::case_when(
          !is.na(l_max_type) & l_max_type == "TL" ~ l_max_fishbase,
          !is.na(l_max_type) & l_max_type %in% c("FL","SL") &
            !is.na(La) & !is.na(Lb) ~ La + Lb * l_max_fishbase,
          TRUE ~ NA_real_
        ),
        # NB: life_histories_summary may NOT have an existing `l_max` column,
        # so don't reference it; compute a new one and expose later.
        l_max_calc = dplyr::coalesce(l_max_from_type, l_max_fishbase),

        # ensure w_inf is present; if missing, derive from L∞ and a,b
        w_inf   = ifelse(!is.na(w_inf), w_inf,
                         ifelse(!is.na(l_inf) & !is.na(a) & !is.na(b),
                                a * (l_inf ^ b), NA_real_)),

        w_mat   = a * (l_mat ^ b),       # weight at maturity
        w_max   = a * (l_max_calc ^ b),  # max weight (from FishBase l_max)
        n       = 0.7,
        p       = 0.7,
        d       = n - 1,
        alpha   = 0.8
    ) %>%
    # join common names
    left_join(
        species_map_lauria,
        by = c("SciName" = "Scientific_name")
    ) %>%
    # overwrite species with common name
    mutate(species = species) %>%
    # keep desired columns
    transmute(
        species,     # common name 
        SciName,     # scientific name 
        w_inf,
        k_vb,
        l_mat,
        l_max = l_max_calc,
        w_max,
        w_mat,
        a, b, t0,
        n, p, d, alpha
    )

# Age at maturity from VB (needs w_inf, k_vb, t0)
sp$age_mat <- mizer:::age_mat_vB(sp)

```

