---
title: "Celtic Sea Operational Model"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)
# core packages
suppressPackageStartupMessages({
  library(mizer)
  library(mizerEcopath)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(yaml)
  library(remotes)
  library(here)
})
```

# Introduction

This document builds the **Celtic Sea Operational Model** in mizer, a functional and reproducible configuration intended for forward projections and scenario analysis - for example, within a Shiny application.

The goal is to produce a model suitable for use in interactive tools and policy exploration, not just for development testing. All parameter choices, data sources, and manual adjustments are recorded so the workflow remains transparent and repeatable.

Manual tuning is required to align the model with empirical data (e.g. survey biomass, growth curves). These subjective adjustments are explicitly, and in future may be reinforced with improved algorithmic matching. 

# Load Data

Load in the csv and rds files which contain species and empirical data. This includes data from DATRAS, STECF, and data from an existing Ecopath model (Lauria et al. 2016, including ongoing updates under Bentley et al.).

```{r}
# Ensure  files exist and pause if any are missing 
req_files <- c(
  here("inst","extdata","Ecopath-Basic estimates.csv"),
  here("inst","extdata","Ecopath-Fishing mortality rates.csv"),
  here("inst","extdata","cs_scaled_landings.csv"),
  here("inst","extdata","cs_survey.csv"),
  here("inst","extdata","Celtic_Sea_Size_at_Age_Data.rds"),
  here("inst","extdata","life_histories.csv"),
  here("inst","extdata","cs_length_maturity_from_age_size_data.csv")
)
stopifnot(all(file.exists(req_files)))

# Ecopath “basic estimates” (contains biomass, production, consumption etc.)
ecopath_params <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

# Ecopath fishing mortality rates by fleet; we aggregate to a total catch proxy
yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"

# Sum across all fleet columns to get a 'Total' F (column indices 3:29 in current file)
# NOTE: This assumes the file layout is stable; if that changes, we will need to revisit this line.
yield$Total <- rowSums(yield[, 3:29], na.rm = TRUE)

# Join to basic estimates to compute total catch (t/km^2/yr) = Total * Biomass
yield <- dplyr::left_join(
  ecopath_params, yield,
  by = c("Group.name" = "Group.name")
) |>
  dplyr::mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
  dplyr::select(Group.name, TotalCatch..t.km..year.)

# ---- Landings (scaled) ----
# Created in inst/Scaling Between Observed Landings and Ecopath.qmd:
# takes landings-by-length from cs_catch.rds and applies a species-specific
# scaling factor so totals match Ecopath catch (Ecopath-Basic estimates.csv +
# Ecopath-Fishing mortality rates.csv). Output is per species × 1 cm length bin,
# with scaled_catch and scaled_biomass in t/km^2 for model calibration.

cs_scaled_landings <- read.csv(here("inst","extdata","cs_scaled_landings.csv"))

# Keep only “total” gear for now (harmonised later when combining)
cs_scaled_landings <- cs_scaled_landings[cs_scaled_landings$gear == "total", ]

cs_scaled_landings <- cs_scaled_landings |>
  # Round down length (floor)
  dplyr::mutate(length = floor(length)) |>
  dplyr::group_by(species, gear, length, Scientific_name) |>
  dplyr::summarise(
    catch   = sum(scaled_catch,   na.rm = TRUE),
    biomass = sum(scaled_biomass, na.rm = TRUE),
    .groups = "drop"
  )

# Add columns expected downstream
cs_scaled_landings$count <- cs_scaled_landings$catch
cs_scaled_landings$dl    <- 1L # L ensures as integer
cs_scaled_landings$gear  <- "landings"

# Reorder to the expected column set/order
cs_scaled_landings <- cs_scaled_landings[, c(
  "species","Scientific_name","length","dl","gear","catch","count","biomass"
)]

# ---- Survey data (DATRAS-derived) ----
# Created in inst/Processing DATRAS Fishing Survey.qmd:
# aggregates DATRAS trawl survey data for the Celtic Sea to kg/km² by
# species × 1 cm length bin, producing scaled catch and scaled biomass.
cs_survey <- read.csv(here("inst","extdata","cs_survey.csv"))

# Aggregate across gears into a single 'survey' gear
cs_survey <- cs_survey |>
  dplyr::group_by(species, Scientific_name, length, dl) |>
  dplyr::summarise(
    catch   = sum(scaled_catch,   na.rm = TRUE),
    biomass = sum(scaled_biomass, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    count = catch,
    gear  = "survey"
  )

# Combined 'catch' table used downstream
catch <- rbind(cs_survey, cs_scaled_landings)

# ---- Age–length data ----
# Created in inst/Processing Size at Age.qmd:
# Species-specific length–age relationships from DATRAS data for growth curves.
age_at_length <- readRDS(here("inst","extdata","Celtic_Sea_Size_at_Age_Data.rds"))

# ---- Life histories (FishBase-derived) ----
# Contains length–weight parameters (a, b), maturity lengths (l_mat),
# and maximum lengths standardised to total length (l_max).
life_histories <- read.csv(here("inst","extdata","life_histories.csv"))

life_histories_summary <- life_histories %>%
  filter(Sex %in% c("Unsexed", "Mixed")) %>%      # keep Unsexed or Mixed
  mutate(Scientific_name = Species) %>%           # duplicate Species to match maturity data
  group_by(Species) %>%
  # Pick row with l_mat if possible, else any row
  slice(
    if (any(!is.na(l_mat))) {
      which(!is.na(l_mat))[1]  # first non-NA
    } else {
      1                        # first row if all NA
    }
  ) %>%
  rename(l_mat1 = l_mat) %>%
  mutate(
    Scientific_name = Species,
    l_max = case_when(
      l_max_type == "TL" ~ l_max_fishbase,
      l_max_type == "FL" ~ La + Lb * l_max_fishbase,
      l_max_type == "SL" ~ La + Lb * l_max_fishbase
    ),
    l_max = ifelse(!is.na(l_max), l_max, l_max_fishbase)
  ) %>%
  ungroup()

# ---- Maturity from processed size-at-age (DATRAS-based) ----
# Created in inst/Processing Size at Age.qmd:
# sets species-specific length at 50% maturity (L50) from DATRAS size-at-age data.
lm <- read.csv(here("inst","extdata","cs_length_maturity_from_age_size_data.csv"))

# Harmonise maturity column names: keep the processed L50 as l_mat2, then merge
lm <- dplyr::rename(lm, l_mat2 = l50)

# Join life histories with maturity; prefer l_mat2 (from processed data), else l_mat1
sp_params <- dplyr::left_join(life_histories_summary, lm, by = "Scientific_name") |>
  dplyr::mutate(
    l_mat = dplyr::if_else(!is.na(l_mat2), l_mat2, l_mat1)
  ) |>
  dplyr::select(-l_mat2, -l_mat1)
```

# Step 1
## 1.1 Species parameters

We first assemble the species parameter table. We merge life-history information (length–weight, maturity length, maximum length) with Ecopath group names and ensure consistent species naming across all inputs.

The goal is to produce a clean, merged object (sp) containing key size- and weight-related parameters for each species, which will be passed to Mizer for baseline parameterisation and later tuning.

Inputs:

life_histories_summary: (FishBase-derived: a, b, l_max, l_mat),
lm: (processed DATRAS maturity; provides l_mat2 already merged),
ecopath_params: (group names/biomass).

```{r}
# Mapping of model "common names" to scientific names used in Lauria/Ecopath
species_map_lauria <- data.frame(
  species = c("Herring","Cod","Haddock","Whiting","Blue whiting","Hake",
              "Monkfish","Horse mackerel","Mackerel","Plaice","Megrim","Sole"),
  Scientific_name = c("Clupea harengus","Gadus morhua","Melanogrammus aeglefinus",
                      "Merlangius merlangus","Micromesistius poutassou","Merluccius merluccius",
                      "Lophius piscatorius","Trachurus trachurus","Scomber scombrus",
                      "Pleuronectes platessa","Lepidorhombus whiffiagonis","Solea solea")
)

# Start from sp_params (built previously by joining life_histories_summary + lm)
sp <- sp_params %>%
    # add computed columns first
    mutate(
        SciName = Species,                # keep scientific name
        t0      = ifelse(is.na(t0), 0, t0),
        l_mat=l_mat,
        l_max=l_max,
        w_mat   = a * (l_mat ^ b),       # weight at maturity
        w_max   = a * (l_max ^ b),       # max weight
        n       = 0.7,
        p       = 0.7,
        d       = n - 1,
        alpha   = 0.8
    ) %>%
        # join common names
    left_join(
        species_map_lauria,
        by = c("SciName" = "Scientific_name")
    ) %>%
    # overwrite species with common name
    mutate(species = species) %>%
    # keep desired columns
    transmute(
        species,     # common name 
        SciName,     # scientific name 
        w_inf,
        k_vb,
        l_mat,
        l_max,
        w_max,
        w_mat,
        a, b, t0,
        n, p, d, alpha
    )

# Age at maturity from VB (needs w_inf, k_vb, t0)
sp$age_mat <- mizer:::age_mat_vB(sp)

```

## 1.2 Add Ecopath Parameters

Next, we link each modeled species to its corresponding Ecopath group(s) (stanza) and import biomass, production, and consumption parameters. Multi-stanza groups (e.g., Cod and Hake) are aggregated.

```{r}
species_to_groups <- list(
    Herring       = "Herring",
    Cod           = c("Cod", "Juvenile cod"),
    Haddock       = c("Haddock","Juvenile haddock"),
    Whiting       = c("Whiting","Juvenile whiting"),
    `Blue whiting`= c("Blue whiting","Juvenile blue whiting"),
    Hake          = c("Hake","Juvenile hake"),
    Monkfish      = c("Monkfish","Juvenile monkfish"),
    `Horse mackerel` = "Horse mackerel",
    Mackerel      = "Mackerel",
    Plaice        = c("Plaice","Juvenile plaice"),
    Megrim        = c("Megrim","Juvenile megrim"),
    Sole          = "Sole"
)

# Use addEcopathParams to validate and add Ecopath parameters to Mizer species.
sp <- addEcopathParams(sp, ecopath_params, species_to_groups)
```

## 1.3 Update w_max from observations 

We adjust each species’ maximum observed length (`l_max`) using the largest length recorded in the scaled landings dataset (`cs_scaled_landings`), which originates from **cs_catch.rds** (observed landings by length, STECF) scaled to match **Ecopath total catches**.

```{r}
observed_max <- cs_scaled_landings %>%
  group_by(species) %>%
  summarise(
    obs_len = if (all(is.na(length + dl))) NA_real_ else max(length + dl, na.rm = TRUE)
  )

sp1 <- sp %>%
    left_join(observed_max, by = "species")%>%
    mutate(
        l_max      = pmax(obs_len, l_max),
        w_max        = a * l_max^b,
    )
```

## 1.4 Build non-interacting mizer model

We next set up a 'non-interacting' mizer model where each species grows according to allometric rules independently of each other species in the assemblage. This essentially generates a series of independent baseline parameters for each species, ready to be matched to observation data. 

The process is:
1. Create allometric size spectra for each species.
2. Add catch data (landings and survey).
3. Assign observed biomasses from Ecopath for matching.
4. Iteratively match growth, biomass, and consumption rates.
5. Launch the shiny tool to facilitate fine hand-tuning alignment with observations. 

This yields a functional, calibrated series of models (wrapped together under the umbrella of a single model) that matches expected growth and biomass, but unable yet to reproduce interspecific assemblage dynamics. 





