---
title: "Celtic Sea Operational Model"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = TRUE,
  fig.align = "center", fig.width = 7, fig.height = 4.5
)
# core packages
suppressPackageStartupMessages({
  library(mizer)
  library(mizerExperimental)  
  library(mizerEcopath)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
  library(yaml)
  library(remotes)
  library(here)
})

# ---- Local package setup ----
# We use a locally edited copy of mizerEcopath (0.1.0.9013) with master mizer (>= 2.5.3.9003).
# The master branch does not export age_mat_vB(), so we provide a small local
# compatibility wrapper (age_mat_vB_compat) and a matching newAllometricParams_compat().

# To load the local mizerEcopath source directly, use:
# devtools::load_all("C:/Documents/York academic/Projects/R projects/mizerEcopath")

```

# Introduction

This document builds the **Celtic Sea Operational Model** in mizer using the **current ** (13.10.2025) main branch (no diffusion). It is set up for scenario projections and interactive exploration (e.g. with a Shiny front end) with an emphasis on reproducibility: all inputs, choices, and manual adjustments are recorded.

**Model scope & key choices**

Fishing data: landings-only gear constructed from updated landings-by-length (scaled to Ecopath totals). Survey gear is not included as separate gear choices are not available for tuning in this mizer release.

Calibration: matchGrowth(), steadySingleSpecies(), matchBiomasses(), and matchCatch() available in this mizer release. Age-at-length data are **not** used in this release - growth is (more loosely) matched to growth curves.

Diet: we import an updated and reformatted Ecopath diet matrix.

# Modified helper functions (for newAllometricParams)

Because the master (non-diffusion) build of mizer does not export the age_mat_vB() function required by newAllometricParams(), a small local compatibility function is added here. It reproduces the expected behaviour for estimating age at maturity from VB parameters and provides a matching newAllometricParams_compat() call that mirrors the original function. This ensures the model can be built on the diffusion-free mizer version without altering the wider workflow.

```{r}
# Local compatibility version mirroring mizer::age_mat_vB()
age_mat_vB_compat <- function(object) {
  ns <- try(getNamespace("mizer"), silent = TRUE)
  if (!inherits(ns, "try-error") && exists("age_mat_vB", envir = ns, inherits = FALSE)) {
    return(get("age_mat_vB", envir = ns, inherits = FALSE)(object))
  }

  if (inherits(object, "MizerParams")) {
    sp <- object@species_params
  } else {
    if (!is.data.frame(object)) {
      stop("The first argument must be either a MizerParams object or a species_params data frame.")
    }
    sp <- validSpeciesParams(object)
  }

  sp <- set_species_param_default(sp, "t0",   0)
  sp <- set_species_param_default(sp, "b",    3)
  sp <- set_species_param_default(sp, "k_vb", NA)
  sp <- set_species_param_default(sp, "w_inf", sp$w_max)

  a_mat <- -log(1 - (sp$w_mat / sp$w_inf)^(1 / sp$b)) / sp$k_vb + sp$t0
  names(a_mat) <- sp$species
  a_mat
}


newAllometricParams_compat <- function(species_params, no_w = 200) {
  sp <- validGivenSpeciesParams(species_params)
  sp <- set_species_param_default(sp, "n", 0.7)
  sp$d <- sp$n - 1
  sp <- set_species_param_default(sp, "alpha", 0.8)
  sp$ks <- 0
  sp$z0 <- 0

  max_w <- max(sp$w_max)
  p <- newMultispeciesParams(
    sp, no_w = no_w, info_level = 0,
    max_w = max_w, w_pp_cutoff = max_w * (1 + 1e-9),
    resource_dynamics = "resource_constant"
  )
  sp <- p@species_params

  interaction_matrix(p)[] <- 0
  sp$interaction_resource <- 0
  sp$h <- Inf
  intake_max(p)[] <- Inf
  species_params(p) <- sp

  ext_encounter(p) <- t(outer(p@w, sp$n, "^"))
  sp <- set_species_param_default(sp, "age_mat", age_mat_vB_compat(sp))
  factor <- age_mat(p) / sp$age_mat
  ext_encounter(p) <- sweep(ext_encounter(p), 1, factor, "*")
  e0 <- getEGrowth(p)[, 1] / w(p)[1]^sp$n[1]

  mu0 <- e0 * (1 + 0.2 - sp$n)
  ext_mort(p) <- sweep(t(outer(p@w, sp$d, "^")), 1, mu0, "*")

  p <- matchBiomasses(p)
  p <- steadySingleSpecies(p, keep = "biomass")
  p <- setBevertonHolt(p, reproduction_level = 0)
  p
}

isAllometric_compat <- function(params, tol = 1e-5) {
  sp <- params@species_params
  sp <- set_species_param_default(sp, "d", sp$n - 1)
  m <- getMort(params) - getFMort(params)
  e <- getEncounter(params)
  for (sp_name in sp$species) {
    spc <- sp[sp_name, ]
    mc <- m[sp_name, ] / params@w^spc$d
    if (anyNA(mc) || any(is.nan(mc)) || any(mc < 0) || all(mc == 0)) stop("The model has invalid mortality rates.")
    if (abs((max(mc) - min(mc)) / min(mc)) > tol) return(FALSE)
    ec <- e[sp_name, ] / params@w^spc$n
    if (anyNA(ec) || any(is.nan(ec)) || any(ec < 0) || all(ec == 0)) stop("The model has invalid encounter rates.")
    if (abs((max(ec) - min(ec)) / min(ec)) > tol) return(FALSE)
  }
  TRUE
}
```

# Load Data

Load in the csv and rds files which contain species and empirical data. This includes data from DATRAS, STECF, and data from an existing Ecopath model (Lauria et al. 2016, including ongoing updates under Bentley et al.).

```{r}
# ---- Check for the required files
req_files <- c(
  here("inst","extdata","Ecopath-Basic estimates.csv"),
  here("inst","extdata","Ecopath-Fishing mortality rates.csv"),
  here("inst","extdata","cs_scaled_landings.csv"),
  here("inst","extdata","life_histories.csv")
)
stopifnot(all(file.exists(req_files)))

# ---- Ecopath basic estimates ----
ecopath_params <- read.csv(here("inst","extdata","Ecopath-Basic estimates.csv"))

# ---- Ecopath fishing mortality 
yield <- read.csv(here("inst","extdata","Ecopath-Fishing mortality rates.csv"))
colnames(yield)[2] <- "Group.name"

# Sum all fleet columns
fleet_cols <- setdiff(seq_len(ncol(yield)), 1:2)
yield$Total <- rowSums(yield[, fleet_cols, drop = FALSE], na.rm = TRUE)

# Join to basic estimates to compute total catch (t/km^2/yr) = Total * Biomass
yield <- left_join(ecopath_params, yield, by = "Group.name") |>
  mutate(TotalCatch..t.km..year. = Total * Biomass..t.km.2.) |>
  select(Group.name, TotalCatch..t.km..year.)

# ---- Landings (scaled; landings-only workflow) ----
# Created in inst/Scaling Between Observed Landings and Ecopath.qmd
cs_scaled_landings <- read.csv(here("inst","extdata","cs_scaled_landings.csv"))

# Keep only total-gear rows; bin to 1 cm; aggregate
cs_scaled_landings <- cs_scaled_landings[cs_scaled_landings$gear == "total", ] |>
  mutate(length = floor(length)) |>
  group_by(species, gear, length, Scientific_name) |>
  summarise(
    catch   = sum(scaled_catch,   na.rm = TRUE),
    biomass = sum(scaled_biomass, na.rm = TRUE),
    .groups = "drop"
  )

# Columns expected downstream
cs_scaled_landings$count <- cs_scaled_landings$catch
cs_scaled_landings$dl    <- 1L # L ensures as integer
cs_scaled_landings$gear  <- "landings"

# Reorder 
cs_scaled_landings <- cs_scaled_landings[, c(
  "species","Scientific_name","length","dl","gear","catch","count","biomass"
)]

# Combined catch table used downstream
catch <- cs_scaled_landings

# ---- Life histories (FishBase-derived; prefer Unsexed/Mixed; no size-at-age merge) ----
# (The parameter set choice is somewhat arbitrary at present)
life_histories <- read.csv(here("inst","extdata","life_histories.csv"))

life_histories_summary <- life_histories %>%
  filter(Sex %in% c("Unsexed", "Mixed")) %>%
  group_by(Species) %>%
  slice(if (any(!is.na(l_mat))) which(!is.na(l_mat))[1] else 1) %>%
  ungroup() %>%
  mutate(Scientific_name = Species)
```

# Step 1
## 1.1 Species parameters

We first assemble the species parameter table. We merge life-history information (length–weight, maturity length, maximum length) with Ecopath group names and ensure consistent species naming across all inputs.

The goal is to produce a clean, merged object (sp) containing key size- and weight-related parameters for each species, which will be passed to Mizer for baseline parameterisation and later tuning.

Inputs:

life_histories_summary: (FishBase-derived: a, b, l_inf, k_vb, t0, l_max, l_mat),
ecopath_params: (group names/biomass).

Note: We no longer use processed DATRAS maturity (lm) or any size-at-age merges in this non-diffusion workflow.

```{r}
# Mapping of model "common names" to scientific names used in Lauria/Ecopath
species_map_lauria <- data.frame(
  species = c("Herring","Cod","Haddock","Whiting","Blue whiting","Hake",
              "Monkfish","Horse mackerel","Mackerel","Plaice","Megrim","Sole"),
  Scientific_name = c("Clupea harengus","Gadus morhua","Melanogrammus aeglefinus",
                      "Merlangius merlangus","Micromesistius poutassou","Merluccius merluccius",
                      "Lophius piscatorius","Trachurus trachurus","Scomber scombrus",
                      "Pleuronectes platessa","Lepidorhombus whiffiagonis","Solea solea")
)

# Start from life_histories_summary (FishBase-derived only)
sp <- life_histories_summary %>%
    # add computed columns first
    mutate(
        SciName = Species,                # keep scientific name
        t0      = ifelse(is.na(t0), 0, t0),
        l_mat   = l_mat,

                # ---- Create l_max ----
        # Convert FL/SL to TL using La + Lb * value where available, else keep TL
        l_max_from_type = case_when(
          !is.na(l_max_type) & l_max_type == "TL" ~ l_max_fishbase,
          !is.na(l_max_type) & l_max_type %in% c("FL","SL") &
            !is.na(La) & !is.na(Lb) ~ La + Lb * l_max_fishbase,
          TRUE ~ NA_real_
        ),
        # NB: life_histories_summary may NOT have an existing `l_max` column,
        # so don't reference it; calculate a new one and expose later.
        l_max_calc = coalesce(l_max_from_type, l_max_fishbase),

        # ensure w_inf is present; if missing, derive from L∞ and a,b
        w_inf   = ifelse(!is.na(w_inf), w_inf,
                         ifelse(!is.na(l_inf) & !is.na(a) & !is.na(b),
                                a * (l_inf ^ b), NA_real_)),

        w_mat   = a * (l_mat ^ b),       # weight at maturity
        w_max   = a * (l_max_calc ^ b),  # max weight (from FishBase l_max)
        n       = 0.7,
        p       = 0.7,
        d       = n - 1,
        alpha   = 0.8
    ) %>%
    # join common names
    left_join(
        species_map_lauria,
        by = c("SciName" = "Scientific_name")
    ) %>%
    # overwrite species with common name
    mutate(species = species) %>%
    # keep desired columns
    transmute(
        species,     # common name 
        SciName,     # scientific name 
        w_inf,
        k_vb,
        l_mat,
        l_max = l_max_calc,
        w_max,
        w_mat,
        a, b, t0,
        n, p, d, alpha
    )

# Age at maturity from VB (needs w_inf, k_vb, t0)
sp$age_mat <- age_mat_vB_compat(sp)

```

## 1.2 Add Ecopath Parameters

Next, we link each modeled species to its corresponding Ecopath group(s) (stanza) and import biomass, production, and consumption parameters. Multi-stanza groups (e.g., Cod and Hake) are aggregated.

```{r}
species_to_groups <- list(
    Herring          = "Herring",
    Cod              = c("Cod", "Juvenile cod"),
    Haddock          = c("Haddock","Juvenile haddock"),
    Whiting          = c("Whiting","Juvenile whiting"),
    `Blue whiting`   = c("Blue whiting","Juvenile blue whiting"),
    Hake             = c("Hake","Juvenile hake"),
    Monkfish         = c("Monkfish","Juvenile monkfish"),
    `Horse mackerel` = "Horse mackerel",
    Mackerel         = "Mackerel",
    Plaice           = c("Plaice","Juvenile plaice"),
    Megrim           = c("Megrim","Juvenile megrim"),
    Sole             = "Sole"
)

# Use addEcopathParams to validate and add parameters from Ecopath to each species.
sp <- addEcopathParams(sp, ecopath_params, species_to_groups)
```

## 1.3 Update w_max from observations 

We adjust each species’ maximum observed length (`l_max`) using the largest length recorded in the scaled landings dataset (`cs_scaled_landings`), which originates from **cs_catch.rds** (observed landings by length, STECF) scaled to match **Ecopath total catches**.

```{r}
## 1.3 Update w_max from observations 
# Use the largest observed length in the (scaled) landings and give a modest buffer.

observed_max <- cs_scaled_landings %>%
  group_by(species) %>%
  summarise(
    obs_len = max(length + dl, na.rm = TRUE),
    .groups = "drop"
  )

sp <- sp %>%
  left_join(observed_max, by = "species") %>%
  mutate(
    # keep l_max if already larger; otherwise raise it up to the observed max
    l_max = if_else(!is.na(obs_len) & obs_len > l_max, obs_len, l_max),
    # set w_max with a 20% buffer 
    len_for_max = coalesce(obs_len, l_max),
    w_max       = 1.2 * a * (len_for_max ^ b),
    # match Lauria’s practice of setting the upper reproductive weight to w_inf
    w_repro_max = w_inf
  ) %>%
  select(-len_for_max, -obs_len)

```

## 1.4 Build non‑interacting mizer model

We next set up a 'non‑interacting' mizer model where each species grows according to allometric rules independently of each other species in the assemblage. This essentially generates a series of independent baseline parameters for each species, ready to be matched to observation data.

The process is:

1. Create allometric size spectra for each species.
2. Add Ecopath total catch targets (for catch matching).
3. Assign observed biomasses from Ecopath for matching.
4. Iteratively match growth, biomass, and consumption rates.

```{r}
# Set up independent (non-interacting) species spectra from sp
model_non_int <- newAllometricParams_compat(sp, no_w = 200)

# Attach Ecopath-derived total catch targets used by matching
# We created `yield` above with TotalCatch..t.km..year.; convert to expected column names
catch_df <- yield %>%
  transmute(
    `Group name` = Group.name,
    `TotalCatch (t/km²/year)` = TotalCatch..t.km..year.
  )

model_non_int <- addEcopathCatchTotal(model_non_int, catch_df)

# ---- Match growth & biomass to observations ----
# Map each species to its adult stanza and join adult biomass from Ecopath.
# We collapse Ecopath to one row per group name to stay robust if the file contains duplicates.
adult_map <- tibble::tibble(
  species      = names(species_to_groups),
  adult_stanza = vapply(species_to_groups, function(x) x[[1]], character(1))
)

adult_biomass <- ecopath_params %>%
  group_by(Group.name) %>%
  summarise(
    biomass_observed = sum(Biomass..t.km.2., na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(adult_stanza = Group.name)

species_params_tbl <- species_params(model_non_int) %>%
  select(-any_of("biomass_observed")) %>%  # drop if present from a previous run
  left_join(adult_map, by = "species") %>%
  left_join(adult_biomass, by = "adult_stanza") %>%
  mutate(
    biomass_observed = coalesce(biomass_observed, 0),
    biomass_cutoff   = w_mat
  ) %>%
  select(-adult_stanza)

species_params(model_non_int) <- species_params_tbl

# Calibration to single-species targets
model_non_int <- model_non_int %>%
  matchGrowth() %>%
  steadySingleSpecies() %>%
  matchBiomasses()


```
## 1.5 Match catch and consumption (landings‑only, not survey)

We now match catches to the landings‑only observations and then align consumption with Ecopath reference values. (Survey and size‑at‑age inputs are not used in this non‑diffusion build.)

```{r}
## 1.5 Match catch and consumption (landings-only)
# - use landings only
# - standardise a single gear label

landings_total <- cs_scaled_landings %>%
  dplyr::mutate(
    # harmonise casing to model names if needed
    species = dplyr::recode(species, "Horse Mackerel" = "Horse mackerel")
  ) %>%
  dplyr::select(species, length, dl, count) %>%
  dplyr::mutate(gear = "total")

model_non_int <- model_non_int %>%
  matchCatch(catch = landings_total, yield_lambda = 1) %>%
  matchConsumption()

```

## 1.6 Add diet and prepare for interactions.
In this step we add diet information to the model, using the updated Ecopath diet matrix (reformatted species × prey table). This enables diet‑based tuning and subsequent activation of  interactions with (matchDiet()).

```{r}
## 1.6 Add diet and prepare for interactions --------------------------------
## Reduce the Ecopath diet to a species×species matrix and set kernels from
## stomach-fit summaries (no joins; align by order like the original CS build).

# Diet (Ecopath -> model species × species)
ecopath_diet <- read.csv(
  here("inst", "extdata", "Ecopath-Diet_updated_reformatted.csv"),
  check.names = FALSE
)

dm <- reduceEcopathDiet(
  species_params = species_params(model_non_int),
  ecopath_diet   = ecopath_diet,
  ecopath_params = ecopath_params
)

# Stomach-fit parameters (keep separate; align by scientific name)
stf <- readRDS(here("inst", "extdata", "stomach_data_fit.rds"))

sp  <- species_params(model_non_int)
idx <- match(sp$SciName, stf$species)  # align order to species_params

lambda <- 2
sp$pred_kernel_type <- "power_law"
sp$kernel_exp <- stf$alpha[idx] + 4/3 - lambda
sp$kernel_l_l <- stf$ll[idx]
sp$kernel_u_l <- stf$ul[idx]
sp$kernel_l_r <- stf$lr[idx]
sp$kernel_u_r <- stf$ur[idx]

species_params(model_non_int) <- sp

# Save a checkpoint
#saveRDS(model_non_int, here("data","model_non_int.rds"))
```

## 1.8 Pre-tuning set up
Sets up elements required before calling tuneEcopath

```{r}
# Directory for checkpoints
chk_dir <- here("data")

# Modify landings data for tuning (must have species, length, dl, catch, gear)
catch_for_tune <- cs_scaled_landings %>%
  mutate(
    species = recode(species, "Horse Mackerel" = "Horse mackerel"),
    gear    = "total"
  ) %>%
  transmute(
    species,
    length,
    dl     = as.integer(dl),
    catch  = as.numeric(count),
    gear
  )

# Empty survey age-at-length stub so the gadget's Survey Age tab does not error [temporary solution due to using new mizerEcopath gadget]
age_at_length_empty <- tibble::tibble(
  species = character(),
  survey  = character(),
  length  = numeric(),
  dl      = integer(),
  age     = integer(),
  count   = numeric()
)
```

## 1.9 Add fishing mortality to allometric background
After matching catches and yield, we can extract F (fishing mortality) in the fitted model, and add it to the allometric external mortality. Tuning can continue from there.

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Add total fishing mortality (F) on top of existing external mortality
# As such, first tune to create a F-tuned model, then extract the parameters from that. 
add_F_to_extmort <- function(p, F) {
  ext_mort(p) <- getExtMort(p) + F
  p
}

## Tune to landings distribution and yield (start with model_non_int and switch to f_tuned_model if partially tuned):
f_tuned_model <- tuneEcopath(
    f_tuned_model,
    controls = c("match","growth",
                 "diffusion",
                 "fishing",
                 "reproduction",
                 "other"),
    catch         = catch_for_tune,
    age_at_length = age_at_length_empty,
    diet          = dm
)

# 1. Estimate F from the tuned model
F_mat <- getFMort(f_tuned_model)

# 2. Add F onto the model’s external mortality
f_tuned_model_MORT <- add_F_to_extmort(f_tuned_model, F_mat)

#saveRDS(f_tuned_model, here("data","f_tuned_model.rds"))
# Load the checkpoint if required - will need to call add_F_to_extmort on this object.
#f_tuned_model <- readRDS(file.path(chk_dir, "f_tuned_model.rds"))

```

## 2 Manual tuning 

Rounds of re-tuning with saved model object. 

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Load the checkpoint saved at the end of 1.6 if required
#f_tuned_model <- readRDS(file.path(chk_dir, "f_tuned_model.rds"))

# Run the tuner - diffusion sliders will appear but have no effect on the model. 
model_non_int_1 <- tuneEcopath(
    f_tuned_model_MORT,
    controls = c("match","growth",
                 "diffusion",
                 "fishing",
                 "reproduction",
                 "other"),
    catch         = catch_for_tune,
    age_at_length = age_at_length_empty,
    diet          = dm
)

# With the set up above, untick catch, press match, to save a state which can be returned to (pressing with catch ticked may return to the original catch configuration, which controls catchability, selectivity, and mortality).

# Optional: save tuned checkpoint [will need to update the tuneEcopath command to use this instead to continue]
# saveRDS(model_non_int_1, file.path(chk_dir, "model_tuned_1.rds"))

```

# 2.1 Further tuning on existing model object

Use this to reopen tuneEcopath on the current model (model_non_int_1) without rebuilding. We explicitly set controls so the UI initialises cleanly (the diffusion tab is shown but inert). The app is fed the scaled landings (catch_for_tune), an empty age–length placeholder (age_at_length_empty), and the Ecopath-reduced diet matrix (dm).

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Area to rework partially tuned model
model_non_int_1 <- tuneEcopath(
    model_non_int_1,
    controls = c("match","growth",
                 "diffusion",
                 "fishing",
                 "reproduction",
                 "other"),
    catch         = catch_for_tune,
    age_at_length = age_at_length_empty,
    diet          = dm
)

# saveRDS(model_non_int_1, file.path(chk_dir, "model_tuned_1.rds"))
```

# Post-hoc helpers
Sections below contain helpers which can be used to make changes to the feeding kernels and observed biomasses of species.

# Post-hoc: Kernel adjustment

Use this section to edits to the predator–prey kernels). After any change, re-run matchGrowth(model_non_int_1) to re-equilibrate growth with the new kernel before reopening tuneEcopath.

```{r, warning=FALSE, message=FALSE, error=FALSE}

model_non_int_1 <- readRDS(file.path(chk_dir, "model_tuned_1.rds"))

# widen_kernel_respecting_order(): expands each interval around its midpoint,
# then writes back with the SAME original order (no swapping).
widen_kernel_respecting_order <- function(p, sp_name,
                                          factor_length = 1.15,
                                          factor_ratio  = 1.15,
                                          delta_exp     = 0) {
  sp <- species_params(p); i <- match(sp_name, sp$species); stopifnot(!is.na(i))

  # helpers: expand [a,b] by factor around midpoint, but return in original order
  expand_keep_order <- function(lo, hi, factor) {
    # keep original order
    orig_lo <- lo; orig_hi <- hi
    a <- min(lo, hi); b <- max(lo, hi)
    mid <- (a + b) / 2
    half <- (b - a) / 2
    half_new <- half * factor
    a2 <- mid - half_new
    b2 <- mid + half_new
    # restore original order
    if (orig_lo <= orig_hi) c(a2, b2) else c(b2, a2)
  }

  old <- sp[i, c("kernel_l_l","kernel_u_l","kernel_l_r","kernel_u_r","kernel_exp"), drop = FALSE]

  # length-ratio window
  len_new <- expand_keep_order(sp$kernel_l_l[i], sp$kernel_u_l[i], factor_length)
  sp$kernel_l_l[i] <- len_new[1]
  sp$kernel_u_l[i] <- len_new[2]

  # mass-ratio window
  rat_new <- expand_keep_order(sp$kernel_l_r[i], sp$kernel_u_r[i], factor_ratio)
  sp$kernel_l_r[i] <- rat_new[1]
  sp$kernel_u_r[i] <- rat_new[2]

  # exponent adjust
  sp$kernel_exp[i] <- sp$kernel_exp[i] + delta_exp

  species_params(p) <- sp

  message(sprintf(
    "%s kernel: L[%g,%g] R[%g,%g] exp=%g  -->  L[%g,%g] R[%g,%g] exp=%g",
    sp_name, old$kernel_l_l, old$kernel_u_l, old$kernel_l_r, old$kernel_u_r, old$kernel_exp,
    sp$kernel_l_l[i], sp$kernel_u_l[i], sp$kernel_l_r[i], sp$kernel_u_r[i], sp$kernel_exp[i]
  ))
  p
}

# Example:  X % widening for Cod, no exponent change
model_non_int_1 <- widen_kernel_respecting_order(model_non_int_1, "Cod",
                                                 factor_length = 1.4,
                                                 factor_ratio  = 1.4,
                                                 delta_exp     = 0)

# Re-align growth before re-opening the tuner
model_non_int_1 <- matchGrowth(model_non_int_1)

```

# Post-hoc: Biomass adjustment

This helper allows quick manual scaling of the observed biomass target for selected species after tuning. It’s intended for fine adjustments — useful when Ecopath totals look implausible. The change affects reproduction equilibrium only and does not alter gear selectivity or fishing mortality. After scaling, the model re-equilibrates with matchBiomasses() and steadySingleSpecies() to maintain balance.

```{r, warning=FALSE, message=FALSE, error=FALSE}
# Post-hoc: multiply observed biomass targets and show a simple before/after
# - factors: named numeric vector (e.g. c("Megrim" = 1.4, "Cod" = 1.2))
#            OR a 2-column data.frame: species, factor
# - reequilibrate: if TRUE, run matchBiomasses() and steadySingleSpecies()
# Returns the updated model and prints a table for changed species.
scale_observed_biomass <- function(p, factors, reequilibrate = TRUE, digits = 3) {
  sp <- species_params(p)
  sp_names <- as.character(sp$species)

  # Normalise `factors` to a named numeric vector
  if (is.data.frame(factors)) {
    if (ncol(factors) < 2) stop("Provide a 2-column data.frame: species, factor")
    fvec <- setNames(as.numeric(factors[[2]]), as.character(factors[[1]]))
  } else if (is.numeric(factors) && !is.null(names(factors))) {
    fvec <- factors
  } else {
    stop("Provide a named numeric vector (names = species) or a 2-col data.frame.")
  }

  # Keep only species that exist in the model
  hit <- intersect(names(fvec), sp_names)
  if (!length(hit)) stop("None of the provided species match species_params$species")

  # Save original targets
  original_target <- setNames(as.numeric(sp$biomass_observed), sp_names)

  # Apply multiplicative scaling to targets
  idx <- match(hit, sp_names)
  sp$biomass_observed[idx] <- sp$biomass_observed[idx] * fvec[hit]
  species_params(p) <- sp

  # Re-equilibrate if requested (by default runs matchBiomasses and steady).
  if (reequilibrate) {
    p <- matchBiomasses(p)
    p <- steadySingleSpecies(p, keep = "biomass")
  }

  # New targets + realised biomasses from the model
  sp_after <- species_params(p)
  new_target <- setNames(as.numeric(sp_after$biomass_observed), as.character(sp_after$species))

  realised_vec <- tryCatch(
    getBiomass(p),
    error = function(e) setNames(rep(NA_real_, length(sp_names)), sp_names)
  )

  # Small, readable comparison for the species you changed
  out <- data.frame(
    species             = hit,
    original_target     = as.numeric(original_target[hit]),
    new_target          = as.numeric(new_target[hit]),
    realised            = as.numeric(realised_vec[hit]),
    realised_to_target  = as.numeric(realised_vec[hit]) /
                          pmax(as.numeric(new_target[hit]), .Machine$double.eps),
    check.names = FALSE
  )

  # Pretty print (optional rounding)
  num_cols <- vapply(out, is.numeric, logical(1))
  out[num_cols] <- lapply(out[num_cols], function(x) signif(x, digits))
  rownames(out) <- NULL
  print(out)
  invisible(p)
}

## Example usage:
# factors <- c("Megrim" = 1.4, "Cod" = 1.2)
# model_non_int_1 <- scale_observed_biomass(model_non_int_1, factors)

factors_cod <- c("Whiting" = 1.5) # Other uses run in console, check with helper below
model_non_int_1 <- scale_observed_biomass(model_non_int_1, factors_cod, reequilibrate = TRUE)

### Compare biomass in model to original to check for any changes:

compare_biomass_to_ecopath <- function(mod) {
  # Accept either an object or a character name
  if (is.character(mod)) mod <- get(mod, envir = parent.frame())

  adult_map <- data.frame(
    species      = names(species_to_groups),
    adult_stanza = vapply(species_to_groups, function(x) x[[1]], character(1))
  )

  ecopath_adult <- ecopath_params %>%
    group_by(Group.name) %>%
    summarise(bio_ecopath = sum(Biomass..t.km.2., na.rm = TRUE), .groups = "drop")

  out <- species_params(mod) %>% as.data.frame() %>%
    select(species, bio_model = biomass_observed) %>%
    left_join(adult_map, by = "species") %>%
    left_join(ecopath_adult, by = c("adult_stanza" = "Group.name")) %>%
    mutate(ratio = bio_model / pmax(bio_ecopath, .Machine$double.eps)) %>%
    arrange(desc(abs(log10(pmax(ratio, .Machine$double.eps))))) %>%
    select(species, bio_ecopath, bio_model, ratio)

  print(out)
  invisible(out)
}

# example:
compare_biomass_to_ecopath(model_non_int_1)

```




